\section{Hamiltonian Cycles}

\begin{definition}

                A \textbf{Hamiltonian Cycle} is a cycle in $G = (V, E)$ that contains every vertex exactly once.
            
\end{definition}

\begin{theorem}

                Let $m, n \ge 2$. An $n \times m$ grid contains a Hamiltonian cycle iff $n \cdot m$ is even.
            
\end{theorem}

\begin{theorem}

                A bipartite graph $G = (A \cup B, E)$ with $|A| \neq |B|$ cannot contain a Hamiltonian cycle.
            
\end{theorem}

\begin{theorem}

\textbf{Dirac's Theorem (1952)}:\\

                Every graph $G = (V, E)$ with $|V| \ge 3$ and minimum degree $\delta(G) \ge |V|/2$ contains a
                Hamiltonian cycle.
            
\end{theorem}

\begin{algorithm}

\textbf{Dynamic Programming for Hamiltonian Cycle}:\\

                Time complexity $O(n^2 2^n)$.
                Define $P_{S, x} = 1$ if there is a $1-x$ path containing exactly the vertices in $S$.
                Recursion:\\

                $$P_{S, x} = \max \{ P_{S \setminus \{x\}, y} \mid y \in N(x) \cap S, y \neq 1 \}$$
            
\end{algorithm}

\subsection{Travelling Salesman Problem (TSP)}

            The travelling salesman problem is the problem of finding a Hamiltonian cycle in a complete graph $K_n$ with
            minimal weight. In other words, the shortes roundtrip/roundtour.

            \begin{theorem}

                The travelling salesman problem is NP-Complete.
            
\end{theorem}

            This is easy to prove by reduction from Hamiltonian Cycle. Assume we have a
            Graph $G$ and want to check if it is Hamiltionian. Now lets extend the graph to be complete and set the
            weight $w_e$ of every edge $e = (u,v) \in G$ to 0 and the weight of all the newly added edges to 1. Now we
            can check if the TSP has a solution with weight 0. If it does, then $G$ has a Hamiltonian cycle, otherwise
            it does not. Therfore TSP has to be NP-Complete as Hamiltionian decision is to.

            \subsubsection{Metric TSP}

            Assume we have a graph $G$ for which the triangle inequality holds, i.e.:
            $$w(u,v) \le w(u,w) + w(w,v) \quad \forall u,v,w \in V$$

            For this problem there is a 2 and even a 1.5 approximation algorithm.
            \begin{algorithm}

\textbf{2-Approximation for Metric TSP}:\\

\begin{itemize}
    \item Determine a minimal-spanning Tree $T$
    \item Double all the edges of $T$ to get an Eulerian multigraph $T'$
    \item Go along that Eulertour and skip every vertex that has already been visited
    \item The resulting tour is a Hamiltonian cycle
\end{itemize}

                This Hamiltionian cycle has length at most $2 \cdot |E(T)| = 2 \cdot (|V|-1)$.

            
\end{algorithm}

