\section{Pipelining}

\begin{definition}
\textbf{Pipelining} overlaps multiple instructions in execution (assembly line logic). Increases \textbf{instruction throughput} but does not decrease latency.
\end{definition}

\begin{theorem}
A $k$-stage pipeline ideal speedup is $k$-fold, but limited by unbalanced stages, sequencing overhead, and hazards.
\end{theorem}

\subsection{The 5-Stage Pipeline}

\begin{definition}
Stages (MIPS): 1. Fetch (IF), 2. Decode (ID), 3. Execute (EX), 4. Memory (MEM), 5. Write-Back (WB).
\textbf{Pipeline Registers} (e.g. IF/ID) pass data and control signals forward between stages.
\end{definition}

\subsection{Hazards}

\begin{definition}
A hazard prevents the next instruction from executing normally:
\begin{itemize}
    \item \textbf{Structural}: Conflict for hardware resources.
    \item \textbf{Data}: Instruction depends on data from a non-completed instruction.
    \item \textbf{Control}: Branch outcome delays fetching correct next instruction.
\end{itemize}
\end{definition}

\begin{definition}
Data dependencies:
\textbf{Read-After-Write (RAW)}: Instruction needs data about to be written. (Only hazard needing hardware fix in 5-stage pipeline).
\end{definition}

\subsection{Resolving Hazards}

\begin{lemma}
\textbf{Forwarding (Bypassing)} routes result from EX or MEM directly to ALU input.
\end{lemma}

\begin{important}
\textbf{Load-Use Hazard} cannot be forwarded. Requires a \textbf{Stall} (freeze PC, insert pipeline bubble).
\end{important}

\begin{remark}
\textbf{Control Hazards} force a pipeline flush on misprediction. Strategies: Static prediction, Dynamic prediction (Branch History Table), Early Resolution (moving compare to Decode).
\end{remark}

\subsection{Advanced Issues}

\begin{definition}
\textbf{Precise Exceptions}: Must stop execution on exception instruction while ensuring prior instructions finish correctly. Unplanned internal changes are \textbf{exceptions}, external are \textbf{interrupts}. Hardware must flush younger instructions and save PC/REG.
\end{definition}

\begin{definition}
\textbf{Out-of-Order Completion}: Multi-cycle units cause instructions to finish out of order, complicating exceptions.
\textbf{Reorder-Buffer (ROB)}: Stores results of instructions as they complete. Allows out-of-order completion but \textbf{in-order retirement} to maintain precise exceptions. Uses a valid bit in REG to point to ROB if not ready.
\end{definition}

\begin{definition}
\textbf{Fine-Grained Multithreading}: Switches thread every clock to hide hazards.
\end{definition}
