\section{Random Variables}

\begin{definition}
    A \textbf{random variable } is a map $X: \Omega \to \mathbb{R}$ where $\Omega$ is the set of all events of a probability space.
\end{definition}

\begin{example}
    The event that the random variable X takes a value less than or equal to 5 is denoted as:
    $$X \leq 5 \coloneqq \{\omega \in \Omega \mid X(\omega) \leq 5\}$$
\end{example}

\begin{definition}
    The \textbf{probability mass function} of a discrete random variable $X$ is defined as:
    $$f_X(x) = P(X=x)$$
\end{definition}

\begin{definition}
    The \textbf{distribution function} of a discrete random variable $X$ is defined as:
    $$F_X(x) = P(X \leq x)$$
\end{definition}

\subsection{Expectation}

\begin{definition}
    Let $X$ be a random variable. The expectation of $X$ is defined as:
    $$
    \mathbb{E}[X] \coloneqq \sum_{\alpha \in W_x} \alpha P(X=\alpha)
    $$
    aslong as the sum is absolutely convergent, else we say the expectation is undefined.
\end{definition}

\begin{lemma}
    For any random variable $X$ we have:
    $$ \mathbb{E}[X] = \sum_{\omega \in \Omega} X(\omega) P(\omega) = \sum_{k=1}^\infty P(X \geq k) - \sum_{k=1}^\infty P(X \leq -k) $$
\end{lemma}

\begin{proof}
    $$\mathbb{E}[X] = \sum_{\alpha \in W_X}\alpha P[X=\alpha] = \sum_{\alpha \in W_X}\alpha\sum_{\omega \in \Omega, X(\omega)=\alpha} P[\omega] = \sum_{\omega \in \Omega} X(\omega) P[\omega]$$
\end{proof}

\begin{theorem}
    Let $X$ be a random variable with values in $W_X \subseteq \mathbb{N}_0$. Then:
    $$\mathbb{E}[X] = \sum_{k=1}^\infty P[X \geq k]$$
\end{theorem}

\begin{proof}
    $$\mathbb{E}[X] = \sum_{k=0}^\infty k P[X=k] = \sum_{k=0}^\infty \sum_{j=1}^k P[X=k] = \sum_{j=1}^\infty \sum_{k=j}^\infty P[X=k] = \sum_{j=1}^\infty P[X \geq j]$$
\end{proof}

\begin{example}
    We throw a coin and we know $P[\text{head}] = p$. We want to know $\mathbb{E}[X]$ with $X \coloneqq \text{number of throws until we get head}$. Using the above theorem we get:
    $$\mathbb{E}[X] = \sum_{k=1}^\infty P[X \geq k] = \sum_{k=1}^\infty (1-p)^{k-1} = \frac{1}{1-(1-p)} = \frac{1}{p}$$
\end{example}

\begin{definition}
    For an event $E \subseteq \Omega$ the corresponding indicator-variable $X_E$ is defined as:
    $$ X_E(\omega) \coloneqq \begin{cases} 1 & \text{if } \omega \in E \\ 0 & \text{if } \omega \notin E \end{cases}$$
\end{definition}

\begin{remark}
    For an indicator variable $X_E$ we have:
    $$\mathbb{E}[X_E] = P[E]$$
\end{remark}

\begin{theorem}
    The linearity of expectation says that for random variables $X_1, \dots, X_n$  and $a_1, \dots a_n, b \in \mathbb{R}$ we have:
    $$X \coloneqq \sum_{i=1}^n a_i X_i + b$$
    $$\mathbb{E}[X] = \sum_{i=1}^n a_i \mathbb{E}[X_i] + b$$
\end{theorem}

\begin{remark}
    We can intuitively describe this as "the expectation of a sum is the sum of the expectations".
\end{remark}

\begin{proof}
    $$\mathbb{E}[X] = \sum_{\omega \in \Omega} X(\omega) P[\omega] = \sum_{\omega \in \Omega} (a_1 X_1(\omega) + \dots + a_n X_n(\omega)) P[\omega]$$
    $$ = a_1 \sum_{\omega \in \Omega} X_1(\omega) P[\omega] + \dots + a_n \sum_{\omega \in \Omega} X_n(\omega) P[\omega] = a_1 \mathbb{E}[X_1] + \dots + a_n \mathbb{E}[X_n]$$
\end{proof}

\begin{example}
    We throw a coin $m$ times. Let $X \coloneqq$ number of partial sequences with threetimes heads. What is the expectation of $X$?
    $$X = X_1 + X_2 + \dots + X_{m-2}$$
    where $X_i$ is the indicator variable for the event that the $i$-th partial sequence is three heads.
    $$\mathbb{E}[X_i] = \frac{1}{8}$$
    By the linearity of the expectation we get:
    $$\mathbb{E}[X] = \sum_{i=1}^{m-2} \mathbb{E}[X_i] = \frac{m-2}{8}$$
\end{example}

\begin{remark}
    Recall Stable sets in a graph is a set of vertices that are not adjacent to each other. We will now use a randomized algorithm to determin a maximal stable set.
\end{remark}

\begin{algorithm}
S[v] = 1/0 with probability p
for all u,v in E:
    if S[u] = 1 and S[v] = 1:
        S[v] = 0
return S
\end{algorithm}

this results in a stable set but what is the expected size of S?

\begin{theorem}
    The algorithm produces a stable set of expected size at least $n*P-m*P^2$
\end{theorem}

\begin{proof}
    Let $X$ be the random variable for the size of the stable set $S$ after the first step. Further let $Y$ be the random variable for the amount of edges we looked at in step 2. Observe
    $$\mid S \mid  \geq X - Y$$
    By the linearity of expectation we get:
    $$\mathbb{E}[\mid S \mid] \geq \mathbb{E}[X] - \mathbb{E}[Y]$$
    We know that $\mathbb{E}[X] = nP$. Further we know that $\mathbb{E}[Y] = mP^2$. Thus we get:
    $$\mathbb{E}[\mid S \mid] \geq nP - mP^2$$
\end{proof}

\subsection{Variance}

\begin{definition}
    For a random variable $X$ with $\mathbb{E}[X] = \mu$ the variance is defined as:
    $$Var[X] \coloneqq \mathbb{E}[(X-\mu)^2]$$
    The standard deviation of $X$ is defined as $\sigma \coloneqq \sqrt{Var[X]}$.
\end{definition}

\begin{example}
    Let $P[X=-10^{10}] = 1/2$ and $P[X=10^{10}] = 1/2$. Then $\mathbb{E}[X] = 0$ and $Var[X] = 10^{20}$ and $\sigma = 10^{10}$.
\end{example}

\begin{theorem}
    For a random variable $X$ we have $Var[X] = \mathbb{E}[X^2] - \mathbb{E}[X]^2$
\end{theorem}

\begin{proof}
    $$Var[X] = \mathbb{E}[(X-\mu)^2] = \mathbb{E}[X^2 - 2\mu X + \mu^2] = \mathbb{E}[X^2] - 2\mu \mathbb{E}[X] + \mu^2 = \mathbb{E}[X^2] - \mathbb{E}[X]^2 $$
\end{proof}

\begin{lemma}
    For a random variable $X$ and $a, b \in \mathbb{R}$ we have $Var[aX+b] = a^2 Var[X]$
\end{lemma}

\begin{proof}
    \begin{enumerate}
        \item $Var[X+b] = \mathbb{E}[(X+b-\mathbb{E}[X+b])^2] = \mathbb{E}[(X+b-(\mathbb{E}[X]+b))^2] = \mathbb{E}[(X-\mathbb{E}[X])^2] = Var[X]$
        \item $Var[aX] = \mathbb{E}[(aX-\mathbb{E}[aX])^2] = \mathbb{E}[(aX-a\mathbb{E}[X])^2] = \mathbb{E}[a^2(X-\mathbb{E}[X])^2] = a^2 \mathbb{E}[(X-\mathbb{E}[X])^2] = a^2 Var[X]$
    \end{enumerate}
\end{proof}

\begin{definition}
    For a random variable $X$ we call $\mathbb{E}[X^k]$ the $k$-th moment of $X$ and $\mathbb{E}[(X-\mathbb{E}[X])^k]$ the $k$-th central moment of $X$.
\end{definition}

\begin{remark}
    Observe that the first moment is the expectation itself. Further observe that the second central moment is the variance.
\end{remark}

\subsection{Multiple Random Variables}

\begin{example}
    Throw a coin and a dice. Let $X$ be the outcome of the coin and $Y$ be the outcome of the dice. What is the expectation of $X+Y$?
\end{example}

\begin{definition}
    Random variables $X_1, X_2, \dots, X_n$ are called independent iff for all $a_1, \dots, a_n \in W_{X_1} \times \dots \times W_{X_n}$ we have:
    $$P[X_1=a_1, \dots, X_n=a_n] = P[X_1=a_1] \dots P[X_n=a_n]$$
\end{definition}

\begin{lemma}
    Observe that events $A_1, \dots, A_n$ are independent iff their indicator variables $X_{A_1}, \dots, X_{A_n}$ are independent.
\end{lemma}

\begin{proof}
    Note that the event "$I_{A_j} = \alpha_j$" is equivalent to the event "$A_j$" if $\alpha_j = 1$ and to the event $\bar{A_j}$ if $\alpha_j = 0$. Same for $A_j^{\alpha_j}$.
    If now $A_1, \dots, A_n$ are independent then for all $\alpha_1, \dots, \alpha_n \in \{0,1\}$ we have:
    $$P[I_{A_1} = \alpha_1, \dots, I_{A_n} = \alpha_n] = P[I_{A_1} = \alpha_1] \dots P[I_{A_n} = \alpha_n]$$
    Using the above note we get:
    $$P[A_1^{\alpha_1} \cap \dots \cap A_n^{\alpha_n}] = P[A_1^{\alpha_1}] \dots P[A_n^{\alpha_n}]$$
\end{proof}

\begin{lemma}
    Let $X_1, \dots, X_n$ be independent random variables and $S_1, \dots, S_n \subseteq \mathbb{R}$ be arbitrary sets then:
    $$P[X_1 \in S_1, \dots, X_n \in S_n] = P[X_1 \in S_1] \dots P[X_n \in S_n]$$
    (Further this implies that for $X_1, \dots, X_n$ independent random variables and $I \subseteq \{1, \dots, n\}$ we have $(X_i)_{i \in I}$ are independent.)
\end{lemma}

\begin{proof}
    $$P[X_1\in S_1, \dots , X_n\in S_n] = \sum_{\alpha_1 \in S_1} \dots \sum_{\alpha_n \in S_n} P[X_1=\alpha_1, \dots, X_n=\alpha_n]$$
    by indepencence of $X_i$ we get:
    $$= \sum_{\alpha_1 \in S_1} \dots \sum_{\alpha_n \in S_n} P[X_1=\alpha_1] \dots P[X_n=\alpha_n]$$
    $$= (\sum_{\alpha_1 \in S_1} P[X_1=\alpha_1]) \dots (\sum_{\alpha_n \in S_n} P[X_n=\alpha_n]) = P[X_1\in S_1] \dots P[X_n\in S_n]$$
\end{proof}

\begin{theorem}
    Let $f_1, \dots, f_n: \mathbb{R} \to \mathbb{R}$ be functions in $\mathbb{R}$ and $X_1, \dots, X_n$ be independent random variables. Then $f_1(X_1), \dots, f_n(X_n)$ are independent too.
\end{theorem}

\begin{proof}
    Let $S_i = \{\beta \mid f_i(\beta) = \alpha_i\}$. Then we have:
    $$P[f_1(X_1) = \alpha_1, \dots, f_n(X_n) = \alpha_n] = P[X_1 \in S_1, \dots, X_n \in S_n]$$
    by the previous lemma we get:
    $$= P[X_1 \in S_1] \dots P[X_n \in S_n] = P[f_1(X_1) = \alpha_1] \dots P[f_n(X_n) = \alpha_n]$$
\end{proof}
    
\subsection{Combined Random Variables}

\begin{theorem}
    For two independent random variable $X, Y$ and $Z = X+Y$ we have:
    $$f_Z(\alpha) = \sum_{\beta \in W_X} f_X(\beta) f_Y(\alpha-\beta)$$
\end{theorem}

\begin{remark}
    This can intuitivaly be seen by looking at every possibility for $X = \beta$: for $X+Y = \alpha$ to hold we need $Y = \alpha - \beta$. Since $X$ and $Y$ are independent we can multiply their probabilities and sum that over all $\beta$
\end{remark}

\begin{theorem}
    For random variables $X_1, \dots , X_n$ we have:
    $$\mathbb{E}[X_1 + \dots + X_n] = \mathbb{E}[X_1] + \dots + \mathbb{E}[X_n]$$
\end{theorem}

\begin{remark}
    \textbf{Calculation Rules} for combined random variables
    \begin{enumerate}
        \item $\mathbb{E}[X+Y] = \mathbb{E}[X] + \mathbb{E}[Y]$
        \item $\mathbb{E}[XY] = \mathbb{E}[X] \mathbb{E}[Y]$ if $X, Y$ are independent
        \item $Var[X+Y] = Var[X] + Var[Y]$ if $X, Y$ are independent
    \end{enumerate}
\end{remark}

\begin{theorem}
    For independent random variablex $X_1, \dots , X_n$ we have:
    $$\mathbb{E}[X_1 \dots X_n] = \mathbb{E}[X_1] \dots \mathbb{E}[X_n]$$
    $$Var[X_1 + \dots + X_n] = Var[X_1] + \dots + Var[X_n]$$
\end{theorem}

\begin{proof}
    We prove for $n = 2$ general case can be proved inductively.
    $$\mathbb{E}[(X+Y)^2] = \mathbb{E}[X^2 + 2XY + Y^2] = \mathbb{E}[X^2] + 2\mathbb{E}[XY] + \mathbb{E}[Y^2]$$
    $$(\mathbb{E}[X+Y])^2 = (\mathbb{E}[X] + \mathbb{E}[Y])^2 = \mathbb{E}[X]^2 + 2\mathbb{E}[X]\mathbb{E}[Y] + \mathbb{E}[Y]^2$$
    By subtracting both equations we get:
    $$\mathbb{E}[(X+Y)^2] - (\mathbb{E}[X+Y])^2 = \mathbb{E}[X^2] - \mathbb{E}[X]^2 + \mathbb{E}[Y^2] - \mathbb{E}[Y]^2$$
    Which is equivalent to $Var[X+Y] = Var[X] + Var[Y]$.
\end{proof}

\begin{important}
    Note that $Var[X \cdot Y] = Var[X] \cdot Var[Y]$ does not hold in general.
\end{important}

\begin{theorem}
    Let $X$ be a random variable. For pariwise disjoint events $A_1, \dots, A_n$ with $A_1 \cup \dots \cup A_n = \Omega$  and $P[A_i] > 0 \forall i$ we have:
    $$\mathbb{E}[X] = \sum_{i=1}^n \mathbb{E}[X \mid A_i] P[A_i]$$
\end{theorem}

\begin{proof}
    $$\mathbb{E}[X] = \sum_{\omega \in \Omega} P[X=\omega] \omega = \sum_{i=1}^n \omega \sum_{\omega \in A_i} P[X=\omega | A_i] P[A_i]$$
    $$ = \sum_{i=1}^n A_i \sum_{\omega \in W_X } \omega P[X=\omega | A_i] = \sum_{i=1}^n \mathbb{E}[X \mid A_i] P[A_i]$$
\end{proof}

\begin{theorem}
    \textbf{The waldsche identity}
    Let $N$ and $X$ be independent random variabeles further let
    $$Z \coloneqq \sum_{i=1}^N X_i$$
    where $X_i$ are independent copies of $X$ then:
    $$\mathbb{E}[Z] = \mathbb{E}[N] \mathbb{E}[X]$$
\end{theorem}

\begin{proof}
    $$\mathbb{E}[Z] = \sum_{k=1}^n \mathbb{E}[Z \mid N=k] P[N=k] = \sum_{k=1}^n \mathbb{E}[X_1 + \dots + X_k] P[N=k]$$
    $$= \sum_{k=1}^n k \mathbb{E}[X] P[N=k] = \mathbb{E}[X] \sum_{k=1}^n k P[N=k] = \mathbb{E}[X] \mathbb{E}[N]$$
\end{proof}

\begin{example}
    Throw a dice where $N$ is the face of the dice we get. Afterwards throw a coin $N$-times and let $Z$ be the number of heads. What is the expectation of $Z$?
    $$\mathbb{E}[Z] = \sum_{k=1}^6 P[N=k] \mathbb{E}[Z \mid N=k] = \sum_{k=1}^6 \frac{1}{6} \frac{k}{2} = \frac{1}{12} \sum_{k=1}^6 k = \frac{1}{12} \frac{6 \cdot 7}{2} = \frac{7}{4}$$
\end{example}

\begin{example}
    Throw a coin until we get heads. Let $N$ be the number of throws. Then throw a coin $N$-times and let $Z$ be the number of heads. What is the expectation of $Z$?
    $$\mathbb{E}[N] = 2 (\text{geometric distribution } p = 1/2)$$
    $$Z = \sum_{i=1}^N X_i \quad \text{where Xi are bernulli distributed with} \quad p = 1/2 \text{ hence } \mathbb{E}[X] = 1/2$$
    $$\mathbb{E}[Z] = \mathbb{E}[N] \mathbb{E}[X] = 2 \cdot \frac{1}{2} = 1$$
\end{example}

\begin{remark}
    The Waldsche identity can be used to analyse the expected runtime of randomized algorithms. Where $N$ is the number of calls of the subroutine and $X$ the runtime of a single call. We can caluclate the total expected runtime using $Z$.
\end{remark}