\section{Branch Prediction}

\begin{remark}
    \textbf{The Goal:} Branch prediction handles control dependencies. It attempts to answer the fundamental question: How do we exploit instruction-level parallelism in the presence of control flow?
\end{remark}

\begin{remark}
    \textbf{The Cost of Misprediction:} Branch mispredictions drastically reduce IPC in deep pipelines. Modern processors therefore target 95\%+ accuracy.
\end{remark}

\begin{definition}
    There are multiple different \textbf{Branch Types}: Conditional (if, while), Unconditional (goto), Indirect (register targets), \& Function calls/returns. \textit{Conditional} and \textit{indirect} are the hardest to predict.
\end{definition}

\begin{remark}
    Control dependencies are handled by:
    \begin{itemize}
        \item \textbf{Stalling} or \textbf{Predicting} the pipeline.
        \item \textbf{Delayed Branching:} Executing adjacent instructions unconditionally.
        \item \textbf{Multi-threading:} Executing other threads.
        \item \textbf{Predicated Execution:} Converting control to data-dependence.
        \item \textbf{Multi-path Execution:} Fetching all branches simultaneously.
    \end{itemize}
\end{remark}

\subsection{Static Branch Prediction}

\begin{definition}
    \textbf{Static Branch Prediction}: The branch outcome is predicted strictly at compile time. Because it is static, it cannot adapt to dynamic changes in branch behavior during runtime.
\end{definition}

\subsubsection*{Basic Static Strategies}
\begin{remark}
    The simplest strategy is to always predict the same outcome (\textbf{Always Taken} or \textbf{Always Not-Taken}). This logic requires minimal hardware but yields low accuracy.
\end{remark}

\begin{definition}
    \textbf{BTFN (Backward Taken, Forward Not-Taken)}: Predicts backward branches as taken and forward branches as not-taken.
\end{definition}
\begin{remark}
    BTFN significantly improves prediction success by intelligently exploiting loop structures. Backward branches are typically loop iterations, which are statistically taken much more often than not.
\end{remark}

\subsubsection*{Advanced Static Strategies}

\begin{definition}
    \textbf{Profile-based Prediction}: Predicts the outcome based on branch behavior statistics gathered from a previous profiling run of the program.
\end{definition}

\begin{definition}
    \textbf{Program-based Prediction}: Uses compiler heuristics (e.g., assuming a check for a negative integer is an error path, hence not-taken).
\end{definition}

\begin{definition}
    \textbf{Programmer-based Prediction}: Relies on explicit pragmas or hints provided by the programmer directly in the source code.
\end{definition}

\begin{codeexample}
    // Pragmas as hints for the branch predictor
    if (likely(x)) {...}
    if (unlikely(error)) {...}
\end{codeexample}

\subsection{Dynamic Branch Prediction}

\begin{definition}
    \textbf{Dynamic Branch Prediction}: The branch outcome is predicted at runtime based on the execution history.
\end{definition}

\subsubsection*{Target vs. Direction Prediction}
\begin{definition}
    \textbf{Branch Target Buffer (BTB)}: A hardware cache that specifically stores the \textit{target addresses} of recently executed branches. When a branch instruction is fetched, the BTB is accessed to provide the target address immediately, avoiding pipeline stalls while waiting for the ALU to compute the target.
\end{definition}

\begin{definition}
    \textbf{Return Address Stack (RAS)}: Dedicated hardware for predicting function returns. It pushes the return address on a \texttt{call} instruction and pops it on a \texttt{ret}. It is highly accurate (>95\% with just an 8-entry stack) for predicting indirect returns.
\end{definition}

\begin{definition}
    \textbf{Indirect Branch Predictor}: Indirect branches map to multiple targets, making standard BTBs inaccurate. These are typically predicted using a dedicated history-based target predictor (e.g., indexing the BTB with the GHR XORed with the PC).
\end{definition}

\subsubsection*{Local History Predictors}
\begin{definition}
    \textbf{Last-Time Predictor}: Predicts the outcome of a branch to be exactly the same as the outcome of the last time that specific branch was executed.
\end{definition}
\begin{remark}
    Last-Time predictors are simple (one bit per branch). They exploit long loops (e.g. 1000 iterations) with high accuracy but perform extremely poorly on short loops (e.g. 2 iterations) or on branches that alternate heavily.
\end{remark}

\begin{definition}
    \textbf{Two-Bit Counter (Bimodal Predictor)}: Prediction only changes after two consecutive mistakes. This introduces \textbf{hysteresis} (via cyclic Strong/Weak states), preventing the prediction from arbitrarily flipping due to a single anomalous outcome.
\end{definition}
\begin{remark}
    Two-Bit Counters significantly improve prediction accuracy compared to Last-Time predictors (yielding around 0.85-0.90 accuracy).
\end{remark}

\subsubsection*{Correlated (Two-Level) Predictors}
\begin{definition}
    \textbf{Two-Level Prediction}: Attempts to correlate the outcome of a branch using history tracking:
    \begin{itemize}
        \item \textbf{Global History Register (GHR)}: Tracks the outcome of \textit{recent different branches} (global context).
        \item \textbf{Local History Register (LHR)}: Tracks the outcome of \textit{the same branch} in its previous iterations.
    \end{itemize}
\end{definition}

\begin{definition}
    \textbf{Gshare Predictor}: XORs the Global History Register (GHR) with the Program Counter (PC) to index the Pattern History Table (PHT), providing branch-specific context and drastically reducing interference.
\end{definition}

\begin{definition}
    \textbf{Agree Predictor} \& \textbf{Gskew Predictor}: Advanced techniques beyond Gshare to reduce interference. The \textbf{Agree Predictor} predicts whether the outcome will \textit{agree} or \textit{disagree} with a statically determined bias bit. The \textbf{Gskew Predictor} uses multiple PHTs with different hash indexing functions and resolves predictions via a majority vote.
\end{definition}

\subsubsection*{Ensemble Predictors}
\begin{important}
    \textbf{Biased Branches \& Branch Filtering}: Filtering heavily biased branches (e.g., 99\% taken) to simpler predictors saves space in history tables for hard-to-predict branches.
\end{important}

\begin{important}
    \textbf{Tournament Predictor}: In many systems, branch prediction is not done using a single predictor. Instead, multiple different predictors run simultaneously, and a meta-predictor chooses whichever predictor has been the most accurate recently. 
\end{important}

\begin{remark}
    \textbf{Notable Historical \& Modern Implementations:}
    \begin{itemize}
        \item \textbf{Intel Pentium Pro}: First commercially successful out-of-order processor, utilizing two-level global prediction.
        \item \textbf{Alpha 21264}: Famously utilized a highly accurate Tournament Predictor.
        \item \textbf{Intel Pentium M}: Introduced specialized Loop and Jump predictors alongside traditional predictors.
        \item \textbf{AMD Zen 2 (and newer)}: Employs a highly accurate hybrid \textbf{Perceptron + TAGE} multi-level branch predictor.
    \end{itemize}
\end{remark}

\subsection{Advanced Predictors}

\begin{definition}
    \textbf{Perceptron Predictor}: Uses a single-layer neural network to predict the outcome of a branch.
\end{definition}

\begin{remark}
    Efficiently implemented as an \textit{adder tree} followed by a sign check. Achieves top-tier accuracy (0.95+) but \textbf{only learns linearly separable functions} (cannot learn XOR-type correlations).
\end{remark}

\begin{definition}
    \textbf{TAGE (TAgged GEometric history length)}: Uses multiple global history registers of varying, geometric lengths to predict the outcome of a branch.
\end{definition}

\begin{remark}
    TAGE predictors represent state-of-the-art branch prediction. They are more complex to implement than perceptron predictors but achieve the highest structural prediction accuracy available (0.97+).
\end{remark}

\subsection{Alternative Control Flow Handling}

\begin{definition}
    \textbf{Delayed Branching}: The $N$ instructions fetched after the branch (delay slots) execute unconditionally. \\
    \textit{Advantage}: Keeps pipeline full. \textit{Disadvantage}: Ties ISA to pipeline depth; hard to fill slots with useful independent instructions.
\end{definition}

\begin{definition}
    \textbf{Predicated Execution (e.g., CMOV)}: Eliminates branches via condition codes. An operation executes but only commits if the predicate is true.\\
    \textit{Advantage}: Eliminates mispredictions, allows larger basic blocks. \textit{Disadvantage}: Wastes energy fetching/executing paths that are discarded.
\end{definition}