\section{Out of Order Execution}

\begin{remark}
    An instruction that uses operands of a previous instruction is called a \textbf{dependent instruction}. If such a preceding instructions result are not yet available, the dependent instruction must wait for the result to be computed, which stalls an \textbf{in-order pipeline}.
\end{remark}

\begin{definition}
    \textbf{Out-of-order execution (OoO)} is a technique that allows a processor to execute instructions in a different order than they appear in the program. This is done to improve performance.
\end{definition}

\begin{remark}
    OoO tolerates longer latencies by executing other independent instructions between dependent ones that would otherwise stall the pipeline. This improves instructions per cycle (IPC).
\end{remark}

\begin{definition}
    The \textbf{dataflow principle} states that the execution order of instructions should be determined by the availability of their operands rather than their program order.
\end{definition}

\begin{remark}
    Typically OoO is only done in the core. Hence instructions are \textbf{fetched and decoded in order} then they are \textbf{executed out of order} and the results are \textbf{written back in order}.
\end{remark}

\subsection{Mechanisms for OoO}

\begin{definition}
    \textbf{Register-Renaming} eliminates false dependencies (WAW and WAR), which occure because of the reuse of registernames, by decoupling the register names from the actual physical storage locations.
\end{definition}

\begin{definition}
    \textbf{Reservation Stations (RS)} are buffers that store instructions that are waiting for their operands to become available.
\end{definition}

\begin{definition}
    \textbf{Tag Broadcast} is a mechanism that allows the results of instructions to be broadcast to all reservation stations that are waiting for those operands by broadcasting identification tags of available results via a bus.
\end{definition}

\subsection{Tomasulo's Algorithm}

\begin{definition}
    \textbf{Tomasulo's Algorithm} is an algorithm for out-of-order execution that uses reservation stations and tag broadcasting to eliminate false dependencies and improve performance.
\end{definition}

\begin{algorithm}
decode(instruction)

if (free reservation station) {

    reservation station = instruction
    replace registers with tags from Register-Alias Table (RAT)

    if (operands ready) {

        execute to functional unit
        broadcast result and update RAT

    } else {
        observe common data bus for tags of operands
    }

} else {
    wait
}
\end{algorithm}

\begin{important}
    The instruction window size (how many instr. can be fetched/decoded and not yet retired) limits the scalability of Tomasulo's algorithm.
\end{important}

\begin{remark}
    Moder processors use the already discussed reorder buffer (ROB) to ensure in order commitment of results.
\end{remark}

\begin{remark}
    The usage of a \textbf{Scheduler} with reservation stations and a \textbf{reorder buffer} can be visualized as a camel with two humps.
\end{remark}

\begin{remark}
    OoO allows for larger latency tolerance and exploitation of Irregular parallelism but comes at the cost of increased complexity and power consumption.
\end{remark}

\subsection{Memory Dependencies}

\begin{example}
    Instruction $i$ writes to memory adress $x$ while the instruction $i+1$ reads from $x$. In out of order execution it can happen that $i+1$ finishes before $i$ which invalidates its result. Hence we need to handle memory dependencies.
\end{example}

\begin{definition}
    \textbf{Memory Disambiguation} is the process of determining whether a load/store is dependent on a previous load/store.
\end{definition}

\begin{definition}
    \textbf{Solutions to memory dependencies:}
    \begin{itemize}
        \item \textbf{Conservative: } Assume all load/stores are dependend and stall the pipeline to ensure correctness.
        \item \textbf{Agressive: } Assume no load/stores are dependent and pay the extra time/work when a false result is generated.
        \item \textbf{Intelligent: } Try and predict whether a load/store is dependent or not in order to minimize the penalty of false results.
    \end{itemize}
\end{definition}

\begin{remark}
    Observe that the intelligent approach is a trade-off between the conservative and aggressive approaches and ensures the best performance.
\end{remark}

\begin{definition}
    \textbf{Load/Store Queue (LSQ)} is a queue that stores load/store results that have not yet been committed to the reorder buffer.
\end{definition}

\begin{definition}
    \textbf{Store-to-Load Forwarding} is a mechanism that allows the result of a store instruction to be forwarded directly to a load instruction that is waiting for that result. This works by searching the LSQ for a matching address and directly copying the data from there.
\end{definition}

