\section{Pipelining}

\subsection{Principles of Pipelining}

\begin{definition}
\textbf{Pipelining} is a microarchitecture implementation technique where \textbf{multiple instructions are overlapped} in execution. It functions like an assembly line, dividing the processing of an instruction into a sequence of stages so that a different instruction can be processed in each stage simultaneously.
\end{definition}

\begin{important}
The core objective of pipelining is to increase \textbf{instruction throughput} (the number of instructions completed per unit of time). While it significantly improves throughput, it generally does not decrease the \textbf{latency} of a single instruction; in fact, the latency may increase slightly due to the overhead of pipeline registers.
\end{important}

\begin{theorem}
In an ideal scenario where all stages have equal delay and there are no dependencies, a pipeline with $k$ stages provides a $k$-fold speedup. However, in reality, the speedup is limited by:
\begin{itemize}
    \item \textbf{Unbalanced stages:} The clock period is determined by the slowest stage.
    \item \textbf{Sequencing overhead:} Pipeline registers add setup time and clock-to-Q delays.
    \item \textbf{Hazards:} Dependencies between instructions that prevent seamless overlapping.
\end{itemize}
\end{theorem}



\subsection{The 5-Stage Instruction Pipeline}

\begin{definition}
A standard RISC pipeline (like MIPS or LC-3) typically consists of five stages:
\begin{enumerate}
    \item \textbf{Instruction Fetch (IF):} Fetch the instruction from memory and update the PC.
    \item \textbf{Instruction Decode (ID):} Decode the instruction and read operands from the register file.
    \item \textbf{Execute (EX):} Perform an arithmetic/logic operation or calculate a memory address.
    \item \textbf{Memory Access (MEM):} Read from or write to the data memory.
    \item \textbf{Write-Back (WB):} Write the result back into the register file.
\end{enumerate}
\end{definition}

\begin{important}
To maintain the state of different instructions in different stages, \textbf{Pipeline Registers} (IF/ID, ID/EX, EX/MEM, MEM/WB) are placed between the stages. These registers carry the data and control signals forward through the pipeline every clock cycle.
\end{important}

\subsection{Pipeline Hazards and Dependencies}

\begin{definition}
A hazard is a situation that prevents the next instruction in the stream from executing in its designated clock cycle.
\begin{itemize}
    \item \textbf{Structural Hazards:} Conflicts for hardware resources (e.g., using a single memory for both instructions and data).
    \item \textbf{Data Hazards:} An instruction depends on the result of a previous instruction that has not yet completed.
    \item \textbf{Control Hazards:} The next address to fetch is unknown because a branch or jump has not been resolved.
\end{itemize}
\end{definition}

\begin{definition}
Data dependencies are primarily of three types:
\begin{itemize}
    \item \textbf{Flow Dependence (RAW):} Instruction $j$ needs data produced by instruction $i$.
    \item \textbf{Anti-Dependence (WAR):} Instruction $j$ writes to a location that instruction $i$ reads.
    \item \textbf{Output Dependence (WAW):} Both instructions write to the same location.
\end{itemize}
In a standard 5-stage in-order pipeline, only RAW hazards require hardware intervention.
\end{definition}

\subsection{Resolving Data Hazards}

\begin{lemma}
\textbf{Forwarding (Bypassing)} resolves many data hazards by routing the result of an instruction directly from the EX or MEM stage to the ALU input of a following instruction, bypassing the register file.
\end{lemma}

\begin{important}
Forwarding cannot resolve a \textbf{Load-Use Hazard} (where a load is immediately followed by an instruction using the loaded data). This requires a \textbf{Stall}. The hardware must:
\begin{enumerate}
    \item Disable the PC and IF/ID register updates (holding instructions in place).
    \item Flush the ID/EX register (inserting a "bubble" or NOP).
\end{enumerate}
\end{important}

\subsection{Control Hazards and Branching}

\begin{definition}
\textbf{Control hazards} arise from the delay in determining the branch outcome. If the pipeline fetches the wrong instructions, it must \textbf{flush} those instructions once the branch is resolved.
\end{definition}

\begin{remark}
Common strategies for handling branches include:
\begin{itemize}
    \item \textbf{Static Prediction:} For example, "Always Predict Not Taken."
    \item \textbf{Dynamic Prediction:} Using a Branch History Table to predict based on past execution.
    \item \textbf{Early Branch Resolution:} Moving the branch comparison to the Decode stage to reduce the misprediction penalty.
\end{itemize}
\end{remark}

\subsection{Advanced Pipeline Issues}

\begin{important}
\textbf{Precise Exceptions:} A pipelined processor must be able to stop execution on an exception-causing instruction while ensuring all prior instructions complete and no subsequent instructions have modified the architectural state. This is challenging because instructions finish at different times.
\end{important}

\begin{lemma}
In systems with multi-cycle functional units (e.g., a fast integer ALU and a slow divider), instructions can finish \textbf{out of order}. This "out-of-order completion" can break the sequential semantics of the ISA if an exception occurs in the middle of a stream.
\end{lemma}

\begin{definition}
\textbf{Fine-Grained Multithreading:} A technique where the processor switches to a different thread every clock cycle. This hides the latency of data and control hazards by ensuring that instructions currently in the pipeline are independent of one another.
\end{definition}


\subsection{Precise Exceptions and Interrupts}

In a pipelined Architecture, not all instructions take the same amount of time in the execution stage. 
So what happens if a previous instruction causes an exception and the next instruction has already been written back to the register file? 
We need to make sure that the state of the processor is consistent with the ISA. 

\begin{definition}
    Unplanned changes in program exection due to internal problems are called \textbf{exceptions} while due to external events they are called \textbf{interrupts}.
\end{definition}

\begin{important}
    When an exception/interrupt occurs all previous instructions should be retired whil no later instructions should be allowed to proceed nor be retired.
\end{important}

Hence when the oldest instruction ready to be retired is detected to have caused an exception we have to:
\begin{itemize}
    \item Ensure architectural state is consistent (REG-File, PC, Memory)
    \item Flush all younger instructions in the pipeline
    \item Save PC and REG
    \item Redirect to exception handler
\end{itemize}

\begin{remark}
    In Single-Cycle machines this is easy to handle as we only have one instruction in the pipeline at a time which ensures sequential execution semantics.
\end{remark}

\begin{remark}
    Note that we could make every operation take the same amount of time in the pipeline, this would ensure sequential execution semantics but would be very inefficient.
\end{remark}

\subsubsection{Reorder-Buffer}

The idea behind a \textbf{reorder-buffer} is that we complete instructions out of order but reorder them before making the results visible to the architectural state.

\begin{definition}
    A \textbf{Reorder-Buffer (ROB)} is a hardware structure that stores the results of instructions as they complete. It allows instructions to be retired in-order, even if they complete out-of-order.
\end{definition}

\begin{lemma}
    Reordering-Buffer uses a valid bit in the REG to indicate whether the result is valid, if not the REG stores a pointer to the ROB entry where the result will be stored once the instruction completes (ROB also uses a valid bit to indicate readiness).
\end{lemma}

This allows us to write out of order to the REG and ensure in order retirement where we can check in-order for any exceptions.



