\section{Sequences and Series}

\subsection{Sequences}

\begin{definition}
    A \textbf{sequence} (Folge) $(a_n)_{n \in \mathbb{N}_0}$ is an infinite list of numbers, mathematically a mapping $f: \mathbb{N}_0 \to X$. It can be given explicitly or recursively  as
    $$a_n = f(n) \quad \text{or} \quad a_n = g(a_{n-1}, a_{n-2}, \dots)$$
\end{definition}

\begin{example}
    A famous example is the Fibonacci sequence defined by $a_0 = 0, a_1 = 1$ and $a_n = a_{n-1} + a_{n-2}$ for $n \ge 2$.
    The first few terms are $0, 1, 1, 2, 3, 5, 8, 13, 21, 34, \dots$
\end{example}

\begin{definition}
    A sequence $(a_n)$ is called \textbf{convergent} with limit $L \in \mathbb{R}$ if:
    $$\forall \epsilon \gt 0 \exists N \gt 0 \quad \text{ s.t. } \quad \forall n \gt N : |a_n - L| \lt \epsilon$$
    Written as $\lim_{n \to \infty} a_n = L$.If no such $L$ exists, the sequence is \textbf{divergent}. Sequences that diverge to $\pm \infty$ are a special case of divergent sequences.
\end{definition}

\begin{theorem}
    The limit of a sequence is unique aslong as it exists.
\end{theorem}

\begin{proof}
    Assume $(a_n)$ converges to $L_1$ and $L_2$.
    Then for any $\epsilon > 0$ there exists $N_1$ and $N_2$ such that:
    $$|a_n - L_1| < \epsilon \quad \forall n > N_1$$
    $$|a_n - L_2| < \epsilon \quad \forall n > N_2$$
    Then for any $n > \max(N_1, N_2)$:
    $$|L_1 - L_2| = |L_1 - a_n + a_n - L_2| \le |L_1 - a_n| + |a_n - L_2| < 2\epsilon$$
    Since this holds for any $\epsilon > 0$, we must have $L_1 = L_2$.
\end{proof}

\begin{theorem}
    A subsequence $(a_{n_k})_{k \in \mathbb{N}_0}$ of a convergent sequence converges to the same limit $L$.
\end{theorem}

\begin{definition}
    An \textbf{accumulation point} (Häufungspunkt) $A \in \mathbb{R}$ of a sequence fulfills:
    $$\forall \epsilon \gt 0 \forall N \in \mathbb{N}_0 \exists n \ge N \text{ such that } |a_n - A| \lt \epsilon$$
\end{definition}

\begin{theorem}
    $A$ is an accumulation point iff there exists a convergent subsequence with limit $A$.
\end{theorem}

\begin{proof}
    Let $(a_n)$ be a sequence and $A$ an accumulation point. We can easily construct a subsequence $(a_{n_k})$ that converges to $A$ by selecting $n_k$ such that $|a_{n_k} - A| < 1/k$ this is given to exist by the definition of an accumulation point.
    Now let $(a_{n_k})$ be a convergent subsequence with limit $A$ then for any $\epsilon > 0$ there exists $K$ such that $|a_{n_k} - A| < \epsilon$ for all $k > K$ and since $n_k \to \infty$ as $k \to \infty$ by definition it follows that $A$ is an accumulation point.
\end{proof}

\begin{theorem}
    Every convergent sequence has exactly one accumulation point (its limit).
\end{theorem}

\begin{proof}
    For $A$ to be an accumulation point of $a_n$ we need that there are indefinitely many $a_n$ that approach $A$. 
    If $a_n$ converges to $L$ then we can intuitively say that that $A$ has to be $L$.
\end{proof}

\begin{theorem}
Limit arithmetic: For $\lim a_n = K$, $\lim b_n = L$ and $C \in \mathbb{R}$:
\begin{itemize}
    \item $\lim (a_n \pm b_n) = K \pm L$
    \item $\lim (C \cdot a_n \cdot b_n) = C \cdot K \cdot L$
    \item $\lim (a_n / b_n) = K / L$ (if $L \neq 0$ and $b_n \neq 0$)
\end{itemize}
\end{theorem}

\begin{theorem}
    \textbf{Sandwich-Theorem:} If $a_n \le b_n \le c_n$ for $n \ge N_0$, and $\lim a_n = \lim c_n = L$, then $\lim b_n = L$.
\end{theorem}

\begin{proof}
    Let $\lim a_n = \lim c_n = L$ and $\lim b_n = M$. From the above arithmetic we can sa that $a_n \leq b_n \leq c_n$ implies $L \leq M \leq L$ so we have that $b_n$ is squeezed between $a_n$ and $c_n$ and thus must converge to $M = L$.
\end{proof}

\begin{remark}
    The sandwich theorem is often used to find upper and lower bounds for a sequence that converge to the same limit and therfore conclude that the original sequence does too.
\end{remark}

\begin{remark}
    Important limits:
    \begin{itemize}
        \item $\lim_{n \to \infty} \frac{\ln n}{n} = 0$
        \item $\lim_{n \to \infty} n^{1/n} = 1$
        \item $\lim_{n \to \infty} x^{1/n} = 1 \quad (x \gt 0)$
        \item $\lim_{n \to \infty} \left(1 + \frac{x}{n}\right)^n = e^x$
        \item $\lim_{n \to \infty} \frac{x^n}{n!} = 0$
    \end{itemize}
\end{remark}

\begin{definition}
    A sequence is
    \begin{itemize}
        \item \textbf{monotonically increasing (decreasing)} if $a_n \le (\geq) a_m$ for $m \gt n$.
        \item \textbf{upper (lower) bounded} if there exists $M$such that $|a_n| \le (\geq) M \quad \forall n \in \mathbb{N}_0$.
    \end{itemize}
\end{definition}

\begin{theorem}
    \textbf{Bolzano-Weierstrass Theorem:} Every bounded sequence has an accumulation point.
\end{theorem}

\begin{theorem}
    \textbf{Monotone Convergence Theorem:} Every bounded monotonic sequence converges to its supremum (infimum).
\end{theorem}

\begin{definition}
    The limit of the supremum is called \textbf{limes superior} and denoted as:
    $$\limsup_{n \to \infty} a_n = \lim_{n \to \infty} \sup_{k \ge n} a_k$$
    The limit of the infimum is called \textbf{limes inferior} and denoted as:
    $$\liminf_{n \to \infty} a_n = \lim_{n \to \infty} \inf_{k \ge n} a_k$$
\end{definition}

\begin{definition}
    A sequence is a \textbf{Cauchy sequence} iff 
    $$\forall \epsilon \gt 0 \exists N \in \mathbb{N} \text{ such that } \forall m, n \ge N : |a_n - a_m| \lt \epsilon$$
\end{definition}

\begin{theorem}
    A sequence of real numbers converges iff it is a cauchy sequence.
\end{theorem}

\begin{proof}
    $\implies$: Let $(a_n)$ converge to $L$. For any $\epsilon \gt 0$, there exists $N$ such that $\forall n \ge N: |a_n - L| \lt \epsilon / 2$.
    Then for any $m, n \ge N$:
    $$|a_n - a_m| = |a_n - L + L - a_m| \le |a_n - L| + |L - a_m| \lt \frac{\epsilon}{2} + \frac{\epsilon}{2} = \epsilon$$
    Thus, $(a_n)$ is a Cauchy sequence.
    
    $\impliedby$: Let $(a_n)$ be a Cauchy sequence. First, $(a_n)$ is bounded: choose $\epsilon = 1$, then there exists $N$ such that $|a_n - a_{N}| \lt 1$ for all $n \ge N$. Thus $|a_n| \lt |a_N| + 1$ for $n \ge N$, meaning $(a_n)$ is bounded by $M = \max(|a_0|, \dots, |a_{N-1}|, |a_N|+1)$.
    By the Bolzano-Weierstrass theorem, the bounded sequence $(a_n)$ has an accumulation point $L$, so there exists a subsequence $(a_{n_k})$ converging to $L$.
    For any $\epsilon \gt 0$, since $(a_n)$ is Cauchy, there exists $N'$ such that $\forall m, n \ge N': |a_n - a_m| \lt \epsilon / 2$.
    Since $a_{n_k} \to L$, choose a $k$ such that $n_k \ge N'$ and $|a_{n_k} - L| \lt \epsilon / 2$.
    Then for any $n \ge N'$:
    $$|a_n - L| \le |a_n - a_{n_k}| + |a_{n_k} - L| \lt \frac{\epsilon}{2} + \frac{\epsilon}{2} = \epsilon$$
    Thus, the sequence converges to $L$.
\end{proof}

\subsection{Series}

\begin{definition}
    A \textbf{series} (Reihe) is an infinite sum $\sum_{i=0}^\infty a_i$ of the terms of a sequence $(a_n)$.
\end{definition}

\begin{remark}
    A series is said to converge iff the sequence of partialsums converges.
\end{remark}

\begin{theorem}
    \textbf{Necessary condition for convergence:} for $\sum_{i=0}^\infty a_i$ to converge we need $\lim_{n \to \infty} a_n = 0$.
\end{theorem}

\begin{remark}
    Some examples of series are:
    \begin{itemize}
        \item \textbf{geometric series}: $\sum_{n=0}^\infty x^n$ converges to $\frac{1}{1-x}$ iff $|x| < 1$
        \item \textbf{harmonic series}: $\sum_{n=1}^\infty \frac{1}{n}$ converges for $s \gt 1$
    \end{itemize}
\end{remark}

\begin{theorem}
        For \textbf{series with non-negative terms} $a_n \ge 0$, the sequence of partial sums is monotonically increasing. \textbf{It converges iff it is bounded.}
\end{theorem}

\begin{theorem}
    \textbf{Comparison test} (Vergleichskriterium): If $c_n \le b_n \le a_n$ for $n \ge N$:
    \begin{itemize}
        \item If $\sum a_n$ converges, then $\sum b_n$ converges (\textbf{Majorantenkriterium}).
        \item If $\sum c_n$ diverges, then $\sum b_n$ diverges (\textbf{Minorantenkriterium}).
    \end{itemize}
\end{theorem}

\begin{example}
    $$\sum_{n=1}^\infty \frac{1}{n!} = 1 + 1 + \frac{1}{2} + \frac{1}{6} + \dots$$
    We can compare this series to the geometric series 
    $$1 +\sum_{n=1}^\infty (\frac{1}{2})^n = 1 + 2 = 3$$
    Therfore accordint to the Majorantenkriterium the series $\sum_{n=1}^\infty \frac{1}{n!}$ converges.
\end{example}

\begin{definition}
    A series is \textbf{absolutely convergent} if $\sum |a_n|$ converges. Absolute convergence implies normal convergence.
    A \textbf{conditionally convergent} series converges, but not absolutely.
\end{definition}

\begin{theorem}
    \textbf{Riemann's rearrangement theorem:}
    Let $\sum a_n$ be a conditionally convergent series. Then for any $x \in \mathbb{R}$, there exists a rearrangement of the series that converges to $x$.
\end{theorem}

\begin{theorem}
    \textbf{Leibniz Criterion}: For an alternating series $\sum (-1)^n a_n$ with a monotonically decreasing null sequence $(a_n)$, the series converges.
\end{theorem}

\begin{proof}
    Let $(a_n)$ be a non-negative, monotonically decreasing sequence converging to $0$. Let $s_n = \sum_{k=0}^n (-1)^k a_k$ be the $n$-th partial sum.
    For the even partial sums $(s_{2n})$, we have 
    $$s_{2n+2} - s_{2n} = -a_{2n+1} + a_{2n+2} \le 0$$ 
    since $a_{2n+2} \le a_{2n+1}$. Thus, $(s_{2n})$ is monotonically decreasing. Similarly for the odd partial sums $(s_{2n+1})$, we have 
    $$s_{2n+3} - s_{2n+1} = a_{2n+2} - a_{2n+3} \ge 0$$
    meaning $(s_{2n+1})$ is monotonically increasing.
    Note that 
    $$s_{2n+1} = s_{2n} - a_{2n+1} \le s_{2n} \le s_0 = a_0$$
    This limits both bounds: $0 \le s_{2n+1} \le s_{2n} \le a_0$.
    By the Monotone Convergence Theorem, both bounded monotonic subsequences converge: $\lim_{n\to\infty} s_{2n} = L_1$ and $\lim_{n\to\infty} s_{2n+1} = L_2$.
    Finally, 
    $$L_1 - L_2 = \lim_{n\to\infty} (s_{2n} - s_{2n+1}) = \lim_{n\to\infty} a_{2n+1} = 0$$
    So $L_1 = L_2 = L$. Since both the even and odd subsequences converge to the same limit $L$, the entire sequence of partial sums $(s_n)$ converges to $L$, thus the series converges.
\end{proof}

\begin{theorem}
\textbf{Cauchy Criterion:} The series $\sum a_n$ converges iff $$\forall \epsilon \gt 0 \exists N \in \mathbb{N} \text{ such that } \forall m, n \ge N : |\sum_{k=n}^m a_k| \lt \epsilon$$
\end{theorem}

\begin{theorem}
    More criterions for convergence:
    \begin{itemize}
        \item \textbf{Root Criterion} (Wurzelkriterium): Let $\rho = \limsup \sqrt[n]{|a_n|}$. If $\rho \lt 1$, abs conv; if $\rho \gt 1$, div.
        \item \textbf{Ratio Criterion} (Quotientenkriterium): Let $\rho = \lim \left| \frac{a_{n+1}}{a_n} \right|$. If $\rho \lt 1$, abs conv; if $\rho \gt 1$, div.
    \end{itemize}
\end{theorem}

\begin{example}
    $$\sum_{n=1}^\infty n^4 e^{-n}$$
    $$\rho = \lim_{n \to \infty} \frac{(n+1)^4 e^{-(n+1)}}{n^4 e^{-n}} = \lim_{n \to \infty} \left( \frac{n+1}{n} \right)^4 e^{-1} = e^{-1} \lt 1$$
    Therfore the series converges.
\end{example}

\begin{definition}
    A \textbf{power series} (Potenzreihe) centered at $a$ has the form 
    $$\sum_{k=0}^\infty c_k(x-a)^k$$
\end{definition}

\begin{theorem}
    Every power series has a \textbf{convergence radius} $R$, with the \textbf{interval of convergence} $(a-R, a+R)$. $R$ is computed by the formula $R = \frac{1}{\limsup |c_n|^{1/n}}$.
\end{theorem}

\begin{example}
    $$\sum_{n=1}^\infty (-1)^{n+1} \frac{1}{n}x^n$$
    $$\rho = \limsup \frac{1}{n^{1/n}} = 1 \implies R = 1$$
    Therfore the series converges for $$x \in (-1, 1)$$.
\end{example}

\begin{remark}
    For $x = 1$ we have the harmonic series, which diverges.
    For $x = -1$ we have the alternating harmonic series, which converges.
\end{remark}
