\section{Integration}

\subsection{Indefinite Integrals and Techniques}

\begin{definition}
    Functions $F: I \to \mathbb{R}$ such that $F'(x) = f(x)$ for all $x \in (a, b)$ are called \textbf{antiderivatives} of $f$.
\end{definition}

\begin{definition}
    The family of all anti-derivatives of a function $f: I \to \mathbb{R}$ for $I \subseteq \mathbb{R}$ an open interval is called the \textbf{indefinite integral} and :=
    $$\int f(x) \, dx$$
    $$= \{ F: I \to \mathbb{R} : F'(x) = f(x) \} = F(x) + C$$
\end{definition}

\begin{theorem}
    Let $f, g: I \to \mathbb{R}$ be integrable functions and let $c \in \mathbb{R}$. Then:
    $$\int (f(x) \pm g(x)) \, dx$$
    $$= \int f(x) \, dx \pm \int g(x) \, dx$$
    $$\int c f(x) \, dx = c \int f(x) \, dx$$
\end{theorem}

\begin{theorem}
    \textbf{Common Antiderivatives}:
    \begin{center}
    \scriptsize
    \setlength{\tabcolsep}{2pt}
    \begin{tabular}{|l|l||l|l|}
    \hline
    $f(x)$ & $\int f(x) dx$ & $f(x)$ & $\int f(x) dx$ \\ \hline
    $x^p$ & $\frac{x^{p+1}}{p+1}$ ($p \ne -1$) & $\frac{1}{x}$ & $\ln|x|$ \\ \hline
    $e^x$ & $e^x$ & $a^x$ & $\frac{a^x}{\ln a}$ \\ \hline
    $\sin x$ & $-\cos x$ & $\cos x$ & $\sin x$ \\ \hline
    $\tan x$ & $-\ln|\cos x|$ & $\frac{1}{\cos^2 x}$ & $\tan x$ \\ \hline
    $\sinh x$ & $\cosh x$ & $\cosh x$ & $\sinh x$ \\ \hline
    $\frac{1}{1+x^2}$ & $\arctan x$ & $\frac{1}{\sqrt{1-x^2}}$ & $\arcsin x$ \\ \hline
    $\frac{1}{\sqrt{x^2+1}}$ & $\text{arsinh } x$ & $\frac{1}{\sqrt{x^2-1}}$ & $\text{arcosh } x$ \\ \hline
    $\ln x$ & $x \ln x - x$ & $\frac{1}{x^2-1}$ & $\frac{1}{2}\ln|\frac{x-1}{x+1}|$ \\ \hline
    \end{tabular}
    \end{center}
\end{theorem}

\begin{remark}
    \textbf{Integration Tricks:}
    \begin{itemize}
        \item \textbf{Partial Fraction Decomposition:} For $\int \frac{P(x)}{Q(x)}dx$. If $\deg P \ge \deg Q$, use polynomial division first.
        \begin{itemize}
            \item Simple real root: $(ax+b) \to \frac{A}{ax+b}$
            \item Repeated real root: $(ax+b)^k \to \frac{A_1}{ax+b} + \frac{A_2}{(ax+b)^2} + \dots + \frac{A_k}{(ax+b)^k}$
            \item Irreducible quadratic: $(ax^2+bx+c) \to \frac{Ax+B}{ax^2+bx+c}$
        \end{itemize}
        \item \textbf{Weierstrass Substitution:} For integrals of $R(\sin x, \cos x)$, let $t = \tan(x/2)$.
        \item \textbf{Square Completion:} For $\sqrt{ax^2+bx+c}$, complete the square to get $\sqrt{u^2 \pm k^2}$.
        \item \textbf{Substitution Patterns:}
        \begin{itemize}
            \item $\sqrt{a^2-x^2} \to x = a \sin \theta$ or $x = a \cos \theta$
            \item $\sqrt{a^2+x^2} \to x = a \tan \theta$ or $x = a \sinh u$
            \item $\sqrt{x^2-a^2} \to x = a \sec \theta$ or $x = a \cosh u$
        \end{itemize}
        \item \textbf{Symmetry:} $\int_{-a}^a f(x)dx = 0$ if $f$ is odd, and $2\int_0^a f(x)dx$ if $f$ is even.
    \end{itemize}
\end{remark}

\begin{theorem}
    \textbf{Integration by Parts}: For differentiable functions $f(x)$ and $g(x)$ we have:
    $$\int f'(x)g(x) \, dx$$
    $$= f(x)g(x) - \int f(x)g'(x) \, dx$$
\end{theorem}

\begin{theorem}
    \textbf{Substitution Rule}: For functions $f(x)$ and $g(x)$ we have:
    $$\int f(g(x))g'(x) \, dx$$
    $$= \int f(u) \, du = F(g(x)) + C$$
    where $F$ is an antiderivative of $f$ and $u = g(x)$.
\end{theorem}

\subsection{The Definite Integral (Riemann Integral)}

\begin{definition}
    Let $U$ and $L$ denote the upper and lower sums of a bounded function $f$. Then $f$ is called \textbf{Riemann-integrable} iff $\inf U = \sup L$. The value of the definite integral $\int_a^b f(x) \, dx$ is then this common value.
\end{definition}

\begin{theorem}
    \textbf{Integrability Conditions}: Any continuous or monotone function on $[a, b]$ is Riemann-integrable.
\end{theorem}

\begin{theorem}
    \textbf{Properties of Riemann Integral}: Let $f, g$ be integrable and $c \in \mathbb{R}$:
    $$\int_a^b (f(x) \pm g(x)) \, dx = \int_a^b f(x) \, dx \pm \int_a^b g(x) \, dx$$
    $$\int_a^b c f(x) \, dx = c \int_a^b f(x) \, dx$$
    $$f(x) \leq g(x) \implies \int_a^b f(x) \, dx \leq \int_a^b g(x) \, dx$$
    $$\left| \int_a^b f(x) \, dx \right| \leq \int_a^b |f(x)| \, dx$$
    $$\int_{a}^{b} f(x) \, dx = -\int_{b}^{a} f(x) \, dx$$
    $$\int_{a}^{c} f(x) \, dx = \int_{a}^{b} f(x) \, dx + \int_{b}^{c} f(x) \, dx$$
\end{theorem}

\begin{theorem}
    \textbf{Mean Value Theorem for Integrals}: Let $f: [a,b] \to \mathbb{R}$ be continuous. Then there exists a $c \in [a,b]$ such that:
    $$\int_{a}^{b} f(x) \, dx = (b-a)f(c)$$
\end{theorem}

\begin{theorem}
    \textbf{Fundamental Theorem of Integral Calculus}: Let $f: [a,b] \to \mathbb{R}$ be continuous. Then for all $C \in \mathbb{R}$, $F(x)$ is an antiderivative of $f$ where:
    $$F(x) = \int_{a}^{x} f(y) \, dy + C$$
\end{theorem}

\subsection{Sequences of Functions and Integrals}

\begin{definition}
    Let $(f_n)_{n \in \mathbb{N}_0}$ be a sequence of functions.
    \textbf{Pointwise Convergence}: It converges pointwise to $f$ if for every $x$, the sequence of real numbers $(f_n(x))$ converges to $f(x)$.
\end{definition}

\begin{definition}
    \textbf{Uniform Convergence}: It converges uniformly to $f$ if for every $\epsilon > 0$, there exists an $N$ such that for all $n \ge N$ and \textbf{for all} $x$ simultaneously, $|f_n(x) - f(x)| < \epsilon$.
\end{definition}

\begin{theorem}
    \textbf{Swapping Limit and Integral}: If a sequence of integrable functions $(f_n)$ converges \textbf{uniformly} to $f$ on $[a,b]$, then the limit function $f$ is also integrable, and:
    $$\int_a^b f(x) \, dx$$
    $$= \lim_{n \to \infty} \int_a^b f_n(x) \, dx$$
\end{theorem}

\subsection{Definite Integration Techniques}

\begin{lemma}
    If $F: [a,b] \to \mathbb{R}$ is continuously differentiable, then:
    $$F(x) = F(a) + \int_{a}^{x} F'(t) \, dt$$
    $$\int_{a}^{b} F'(t) \, dt = F(b) - F(a) = \Big[ F(t) \Big]_a^b$$
\end{lemma}

\begin{lemma}
    \textbf{Substitution for Definite Integrals}: Let $f$ be continuous and $g$ continuously differentiable.
    $$\int_{a}^{b} f(g(x))g'(x) \, dx$$
    $$= \int_{g(a)}^{g(b)} f(u) \, du$$
    \textbf{Version 2}: If $f'(x) \neq 0$ and $f^{-1}$ is the inverse of $f$:
    $$\int_a^b g(f(x)) \, dx$$
    $$= \int_{f(a)}^{f(b)} g(y) (f^{-1})'(y) \, dy$$
\end{lemma}

\begin{lemma}
    \textbf{Integration by Parts for Definite Integrals}: Let $f, g$ be continuously differentiable.
    $$\int_{a}^{b} f'(x)g(x) \, dx$$
    $$= \Big[ f(x)g(x) \Big]_{a}^{b} - \int_{a}^{b} f(x)g'(x) \, dx$$
\end{lemma}
\subsection{Exam Strategy: Leibniz's Rule}
\begin{gather*}
    \frac{d}{dx} \int_{a(x)}^{b(x)} f(t,x) \, dt = f(b(x),x)b'(x) \\
    - f(a(x),x)a'(x) + \int_{a(x)}^{b(x)} \frac{\partial f}{\partial x} \, dt
\end{gather*}

\subsection{Geometric Applications}
\begin{itemize}
    \item \textbf{Area between curves:} $A = \int_a^b |f(x) - g(x)| dx$.
    \item \textbf{Rotational Volume (around x-axis):} $V = \pi \int_a^b (f(x))^2 dx$.
    \item \textbf{Rotational Volume (around y-axis):} $V = 2\pi \int_a^b x f(x) dx$.
    \item \textbf{Arc Length:} $L = \int_a^b \sqrt{1 + (f'(x))^2} dx$.
\end{itemize}

\subsection{Convergence of Improper Integrals}
\begin{itemize}
    \item \textbf{p-test (to infinity):} $\int_1^\infty \frac{1}{x^p} dx$ converges if $p > 1$ (diverges if $p \le 1$).
    \item \textbf{p-test (to zero):} $\int_0^1 \frac{1}{x^p} dx$ converges if $p < 1$ (diverges if $p \ge 1$).
    \item \textbf{Comparison Test:} If $0 \le f(x) \le g(x)$, then $\int g$ conv $\implies \int f$ conv.
\end{itemize}
