\section{Parallel Algorithms}

\subsection{Parallel Patterns}

\begin{definition}

                A \textbf{reduction} is a parallel operation that produces a single answer from a collection via an
                associative operator using for example D\&C. The order
                of operations does not matter.
            
\end{definition}

\begin{example}

                Max, Min, Sum, Product of an array.
            
\end{example}

\begin{definition}

                A standard \textbf{map} applies a function to each elements of a collection producing an output of the
                same size.
            
\end{definition}

\begin{example}

                Vector addition can be done using a map pattern. We split the vectors into dim(v)/2 recursively
                until we
                have dim(v)=1 then sum the vectors and write the result directly into the result vector.
            
\end{example}

\begin{definition}

                A \textbf{stencil} is an extension of the map pattern that applies a function to each element of a
                collection producing an output of the same size. The function uses the element itself and its
                neighbours.
            
\end{definition}

\begin{example}

                A smoothing kernel (ex. 3x3) that calculates the average value is sled over an image to smooth/blur
                the
                image.
            
\end{example}

\subsection{Parallel Algorithms}

\subsubsection{Prefix-Sum}

\begin{example}

                Prefix-Sum (calculating input[1] + input[2] + ... + input[n]) seams to be unparallisable. But infact
                it
                is!
            
\end{example}

\begin{algorithm}

                upTask(); //divide problem and build tree with sums of subproblems
                downTask(); //fill in the correction values (left_sum + parent_correction)
            
\end{algorithm}

\begin{theorem}

                The span of the prefix sum algorithm is $O(\log n)$. This leads to a parallelism of $\frac{n}{\log n}$.
            
\end{theorem}

\subsubsection{Pack}

            Another application of this problem is the \textbf{Pack} algorithm which removes all elements from an array
            that satisfy a certain condition and packs them into a seperate output.

            \begin{algorithm}

                applyMap(); //applies a map to the input\\

                //results in a binary vector [1, 0, 1, 1, 0, ...]\\

                parallelPrefixSum(); //calculates the prefix sum of the binary vector\\

                //results in [1, 1, 2, 3, 3, ...]\\

                applyPackMap(); //applies another map to the result of the prefix sum\\

                //results in [0, 2, 3, ...] (indices of elements to write to output)
            
\end{algorithm}

\subsubsection{Quicksort}

            We can reduce the limiting factor of the problem division from a span of $O(n)$ to a span of $O(\log n)$
            using the pack-pattern we have seen before.

            \begin{algorithm}

                applyMaps(); //apply maps for greater and smaller than pivot\\

                parallelPrefixSum(); //count how many smallerthan/greatherthan elements come bofore\\

                applyMap(); //index = numLessThan || numTotalLessThan + numGreaterThan
            
\end{algorithm}

            We can do this by using two pack-patterns to partition the data. This leads to a partitioning of $O(\log n)$
            levels times $O(\log n)$ for prefixSum hence $O(\log^2 n)$ total span. This means a parallelism of $\frac{n
            \log n}{\log^2 n} = \frac{n}{\log n}$.

        