\section{Threads and Processes}

\begin{definition}
    \textbf{Process: } A program executing inside an OS (operating-system). Processes share cpu but have their own memory space.
\end{definition}

\begin{definition}
The \textbf{context} of a process is the set of utilities it uses and or operates on. That is:
\begin{itemize}
    \item \textbf{Hardware context: }Instruction counter, Registers, Execution State, Flags ...
    \item \textbf{Memory context: }Virtual Memory, Heap-Space, Stack ...
    \item \textbf{OS-level context:} Process ID, open files ...
\end{itemize}
\end{definition}

The \textbf{Process Lifecycle} is managed by the OS and can be in one of the following states:
\begin{itemize}
    \item \textbf{NEW / CREATED:} The process is being created and loaded from disk into memory.
    \item \textbf{READY:} The process is in main memory waiting to be assigned to the CPU by the scheduler.
    \item \textbf{RUNNING:} The process is actively executing instructions on the CPU.
    \item \textbf{BLOCKED / WAITING:} The process cannot continue executing and is waiting for an event (like I/O access) to complete.
    \item \textbf{TERMINATED:} The process has finished execution and its resources are reclaimed.
\end{itemize}

\begin{definition}
\textbf{Context Switch:} The process of saving the context of a process and loading the context of another process.
\end{definition}

\begin{definition}
\textbf{Process Control Block (PCB):} A table-entry storing relevant process information such as Program counter, Registers, Adress Space, Open files ...
\end{definition}

\subsection{Parallelism}

\begin{definition}

\textbf{Parallelism } is when multiple processes execute simultaneously on diffrent CPU-Cores.
            
\end{definition}

\begin{definition}

\textbf{Concurrency } is when multiple processes execute interleaved on a single CPU-Core.
            
\end{definition}

\begin{remark}
$$\text{Parallelism} \implies \text{Concurrency}$$
$$\text{Concurrency} \not\implies \text{Parallelism}$$
\end{remark}

\subsection{Multithreading}

\begin{definition}
A thread is an independent sequence of exection running in the same OS-process.     
\end{definition}

\begin{remark}
While processes can only communicate through main-memory and have diffrent memory spaces, threads share
the same adress space. This leads to faster and more efficient context-switches as adressspace and PCB loading is not necessary.
\end{remark}

There are diffrent levels of threads:
\begin{itemize}
    \item \textbf{User-Level Threads: } Threads managed by the application using a thread library, ex. Java-Threads managed by JVM
    \item \textbf{Kernel-Level Threads: }Threads managed by to OS, which can schedule them on diffrent cores.
    \item \textbf{CPU-Level Threads:} Threads managed by the CPU, which can schedule them on the physical cores.
\end{itemize}

There are two ways to create a thread in java.

\begin{codeexample}
class MyThread extends Thread { ...
    public void run() {
        ...
    }
}
MyThread thread = new MyThread();
thread.start(); // calls MyThread.run()
\end{codeexample}

or using a runnable, which is recomended to do:

\begin{codeexample}
class MyRunnable implements Runnable { ...
    public void run() {
        ...
    }
}
MyRunnable runnable = new MyRunnable();
Thread thread = new Thread(runnable);
thread.start(); // calls MyRunnable.run()
\end{codeexample}

In its lifecylce a thread can have the follwoing states:
\begin{itemize}
    \item \textbf{NEW: }The thread has been created but not yet started.
    \item \textbf{RUNNABLE: }The thread is ready to run and waiting for the OS to schedule it.
    \item \textbf{RUNNING: }The thread is currently executing.
    \item \textbf{BLOCKED: }The thread is waiting for access to I/O devices
    \item \textbf{WAITING: }The thread is waiting to be notified
    \item \textbf{TERMINATED: }The thread has finished its execution
\end{itemize}

\begin{remark}
Every Thread has an \textbf{ID, Name, Priority and State} which might be usfull for certain tasks.            
\end{remark}

To wait for a thread to finish, we can use the .join() method on the thread we want to wait on. This lets
the main-thread sleep until the joined thread terminates and wakes it up again.

\begin{codeexample}
Thread t = new MyThread();
t.start(); //start t
t.join(); //sleep until t is finished
continueWork(); // this is only called after t finishes
\end{codeexample}

\begin{remark}
.join() creates overhead, for small thread-exection-time this is ineffiecient.
\end{remark}

\begin{important}
If a worker-thread throws an exception the main-thread \textbf{will be woken up, but it will not know of the
error}. We have to set an \textbf{UncaughtExeptionHandler using a try-catch block} when joining threads. For threads that
run indefinetly we use a \textbf{timer task} that interupts threads.
\end{important}

\begin{codeexample}
MyTimerThread extends Thread {
    run(Thread thread, int time) {
        timer.sleep(time);
        thread.interupt();
    }
}

Thread t = new MyThread();
Thread timer = new MyTimerThread();
t.start();
timer.start(t, 1000);
try {
    t.join();
} catch (UncaughtExeptionHandler) {
    System.out.println("Error in thread t");
}
continueWork();
\end{codeexample}

